---
title: 'SummarizedExperiment object'
author:
- name: Lindsay Rutter
date: '`r Sys.Date()`'
package: bigPint
bibliography: bigPint.bib
output:
  BiocStyle::html_document:
    toc_float: true
    tidy: TRUE
vignette: >
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{"Data metrics object"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{bigPint}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

## SummarizedExperiment object

Researchers who prefer using the `SummarizedExperiment` class can feed their data and associated metrics into the `bigPint` methods using one `SummarizedExperiment` object (instead of both a `data` and `dataMetrics` object). We demonstrate here how to create a `SummarizedExperiment` object. In the remaining articles of the vignette, the example code uses the `data` and `dataMetrics` objects as example input. However, at the bottom of each vignette, we also include the corresponding example code that uses intead the `SummarizedExperiment` object to create the exact same plots.

____________________________________________________________________________________

## Example: two treatments

As was shown in the article [Data object](https://lindsayrutter.github.io/bigPint/articles/data.html), the `data` object called `soybean_ir_sub` contained 5,604 genes and two treatment groups, N and P [@soybeanIR]. We can create a `SummarizedExperiment` object that combines aspects of the `data` object and `dataMetrics` object. We start by preparing the `data` and `dataMetrics` objects.

```{r, eval=TRUE, include=TRUE, message=FALSE}
library(bigPint)
library(edgeR)
library(data.table)
library(dplyr)
library(DelayedArray)
library(SummarizedExperiment)

data(soybean_ir_sub)
data = soybean_ir_sub
rownames(data) = data[,1]

y = DGEList(counts=data[,-1])
group = c(1,1,1,2,2,2)
y = DGEList(counts=y, group=group)
Group = factor(c(rep("N",3), rep("P",3)))
design <- model.matrix(~0+Group, data=y$samples)
colnames(design) <- levels(Group)
y <- estimateDisp(y, design)
fit <- glmFit(y, design)

dataMetrics <- data.frame(ID = rownames(data))
for (i in 1:(ncol(fit)-1)){
    for (j in (i+1):ncol(fit)){
        contrast=rep(0,ncol(fit))
        contrast[i]=1
        contrast[j]=-1
        lrt <- glmLRT(fit, contrast=contrast)
        lrt <- topTags(lrt, n = nrow(y[[1]]))[[1]]
        
        setDT(lrt, keep.rownames = TRUE)[]
        lrt <- as.data.frame(lrt)
        colnames(lrt) <- paste0(colnames(fit)[i], "_", colnames(fit)[j], ".", colnames(lrt))
        colnames(lrt)[1] = "ID"
        
        reorderID <- full_join(data, lrt, by = "ID")
        dataMetrics <- cbind(dataMetrics, reorderID[,-c(1:ncol(data))])
    }
}
```

Now, we can tranform our `data` object into a `DelayedMatrix` class object and then input both our `data` and `dataMetrics` objects combined into a `SummarizedExperiment` class object. We can verify that the `assay()` and `rowData()` methods work for accessing our `SummarizedExperiment` object.

```{r, eval=TRUE, include=TRUE, message=FALSE}
dataMetrics$ID <- as.character(dataMetrics$ID)
data = data[,-1]
data <- DelayedArray(data)
se_soybean_ir_sub <- SummarizedExperiment(assays = data, rowData = dataMetrics)
assay(se_soybean_ir_sub)
rowData(se_soybean_ir_sub)
```

____________________________________________________________________________________

## Example: three treatments

Similarly, as was shown in the data page, the `data` object called `soybean_cn_sub` contained 7,332 genes and three treatment groups, S1, S2, and S3 [@brown2015developmental]. We can create a `SummarizedExperiment` object that combines aspects of the `data` object and `dataMetrics` object. We start by preparing the `data` and `dataMetrics` objects.

```{r, eval=TRUE, include=TRUE, message=FALSE}
data(soybean_cn_sub)
data = soybean_cn_sub
rownames(data) = data[,1]

y = DGEList(counts=data[,-1])
group = c(1,1,1,2,2,2,3,3,3)
y = DGEList(counts=y, group=group)
Group = factor(c(rep("S1",3), rep("S2",3), rep("S3",3)))
design <- model.matrix(~0+Group, data=y$samples)
colnames(design) <- levels(Group)
y <- estimateDisp(y, design)
fit <- glmFit(y, design)

dataMetrics <- data.frame(ID = rownames(data))
for (i in 1:(ncol(fit)-1)){
    for (j in (i+1):ncol(fit)){
        contrast=rep(0,ncol(fit))
        contrast[i]=1
        contrast[j]=-1
        lrt <- glmLRT(fit, contrast=contrast)
        lrt <- topTags(lrt, n = nrow(y[[1]]))[[1]]
        
        setDT(lrt, keep.rownames = TRUE)[]
        lrt <- as.data.frame(lrt)
        colnames(lrt) <- paste0(colnames(fit)[i], "_", colnames(fit)[j], ".", colnames(lrt))
        colnames(lrt)[1] = "ID"
        
        reorderID <- full_join(data, lrt, by = "ID")
        dataMetrics <- cbind(dataMetrics, reorderID[,-c(1:ncol(data))])
    }
}
```

Now, we can tranform our `data` object into a `DelayedMatrix` class object and then input both our `data` and `dataMetrics` objects combined into a `SummarizedExperiment` class object. We can verify that the `assay()` and `rowData()` methods work for accessing our `SummarizedExperiment` object.

```{r, eval=TRUE, include=TRUE, message=FALSE}
dataMetrics$ID <- as.character(dataMetrics$ID)
data = data[,-1]
data <- DelayedArray(data)
se_soybean_cn_sub <- SummarizedExperiment(assays = data, rowData = dataMetrics)
assay(se_soybean_cn_sub)
rowData(se_soybean_cn_sub)
```

____________________________________________________________________________________

## References
